\documentclass[12pt]{article}
\usepackage[width=13cm,height=22cm,centering]{geometry}
\pagestyle{empty}

\begin{document}

\noindent
CATCH A WINE AND RIDE IT, DUDE\newline
Nicholas Boer\newline
June 27, 2001

\bigskip
\noindent
SURE IT CAN be corked. Still, wine can stir passions with the force of an ocean. Take the bottle of Jordan cabernet I unleashed on my 21st birthday. I had quaffed it with my girlfriend at Plantation Gardens restaurant on the island of Kauai. It stirred us profoundly, but that was nothing compared to what happened to Moan.

\bigskip
\noindent
For a month, I had been a cook at Plantation Gardens, and Moan had become an instant friend. He was cooking that night and came out to wish me a happy birthday. I offered him a glass of wine. That's when the tsunami hit. If you had seen his eyes as he mimicked my swirling and smelling of the luscious cab, you would have known the surf was up. Those eyes were as intense and entranced as they would be riding a 20-foot wave. This wine was tubular.

\bigskip
\noindent
Mark got his nickname from his exaggerated moaning after a sunrise-to-sunset day of surfing. That love for surfing was part of the reason I had to fire him soon after I became sous chef. Moan, like much of the wait staff, tended to have car trouble or a sudden ear infection whenever the surf was raging.

\bigskip
\noindent
But Moan took it well; we remained close friends through it all. He soon got a job as a waiter. And, amazingly, became the sommelier at one of the island's finest restaurants. Amazing, because that glass of Jordan was the first red wine Moan had ever tasted.

\bigskip
\noindent
The journey from surfer to sommelier took a couple of years. At first, he wanted to relive the experience and would only drink Jordan. But he soon risked cash on other cabs or would try something novel I had gotten ahold of. He had a nose for wine and began sniffing out other varietals, eventually embracing even whites.

\bigskip
\noindent
We found our palates to be remarkably alike. It was rare that one of us would like a bottle that the other found disagreeable. We learned and drank and talked about wine. But we never used approved wine jargon. After a first taste of a mediocre wine, one of us would inevitably say ``It's great,'' and the other would finish ``for cooking.''

\bigskip
\noindent
Even with a \$50 disaster, we would remain upbeat. Breathing in the aroma from oversized goblets, we would look at each other and simultaneously let out a scratchy ``meow.'' Code for ``It smells like cat food.''

\bigskip
\noindent
In the early '80s, there were only a couple of shops where you could buy good wine on Kauai. But sometimes we would get lucky. Moan once found an old bottle of Mondavi reserve cabernet for a few bucks in Bucky's liquor store. It was lying in the refrigerator. Probably had been for years.

\bigskip
\noindent
Moan's knowledge of wine soon outgrew my own, and his passion became infectious. As a waiter, he could get so excited about a particular bottle that every one of his tables would order it. He wasn't pushy. He wasn't stuffy. He just loved wine and wanted to share a discovery.

\bigskip
\noindent
He was overjoyed when he became a sommelier. He could pour and chat about wine all night. And he could order anything he wanted for his list. A list that credited me with a ``Merci, Nicholas'' on the bottom.

\bigskip
\noindent
Trouble was, he couldn't keep his hands off the Bordeaux. He would go to the cellar and stroke the label of a \$200 claret. His throat would go dry. For him, it was like watching the curling surf from a hot shore.

\bigskip
\noindent
He uncorked it. Before long, the owner took note of the rising wine costs and Moan was out on the street or, more specifically, out on the ocean.

\bigskip
\noindent
I had moved back to the mainland by the time Moan had become a sommelier; I found out about his fate through phone calls and a mailed copy of his wine list. After getting married, Moan moved back to the mainland, too, went to chiropractor school and tried to set aside his two liquid passions for a more sensible existence. It sounded like the right thing to do, but when I talked to him, he would sound restless.

\bigskip
\noindent
We lost track of each other for a while. But I got a phone call from him not long ago. He was back on Kauai, broke and single. From his flat, he could see the sea, and he described the view to me between sips of wine. It wasn't Jordan, and he hadn't surfed that day no waves.

\bigskip
\noindent
Moan and the ocean were calm that day. But who knew what might pop up tomorrow.

\end{document}
