\documentclass[12pt]{article}
\usepackage[margin=2cm]{geometry}
\usepackage{amsmath}

\newcommand\BNM{B_{nm}} % absorption
\newcommand\BMN{B_{mn}} % induced emission
\newcommand\AMN{A_{mn}} % spontaneous emission
\newcommand\RHO{\rho(\nu,T)}

\newcommand\ABSORPTION{\substack{\\[1ex] \text{absorption}}}
\newcommand\INDUCED{\substack{\\[1ex] \text{induced}\\ \text{emission}}}
\newcommand\SPONTANEOUS{\substack{\\[1ex] \text{spontaneous}\\ \text{emission}}}

\begin{document}

\noindent
In his 1917 paper, ``On the Quantum Theory of Radiation,''
Einstein uses the following argument to derive Planck's law.
%The argument requires the existence of induced emission, a process which was a theoretical discovery by Einstein.
%Prior to Einstein, no one was aware that induced emission existed.

\bigskip
\noindent
Consider a gas at temperature $T$.
Let $N$ be the number of molecules in the gas
and let $N_n$ be the number of molecules with internal energy $E_n$.
By the Maxwell-Boltzmann distribution we have
\begin{equation*}
\frac{N_n}{N}=p_n\exp\left(-\frac{E_n}{kT}\right)
\tag{1}
\end{equation*}

\noindent
where $k$ is Boltzmann's constant.
The coefficient $p_n$ is a statistical weighting factor that does not depend on $T$.

\bigskip
\noindent
Let us now consider the processes by which an atom or molecule transitions between energy levels.
The processes are absorption, induced emission, and spontaneous emission.
Let $E_m$ and $E_n$ be energy levels such that $E_m>E_n$.
Let $N_{m\rightarrow n}/\Delta t$ be the number of atoms or molecules that transition from energy level $E_m$ to $E_n$ in time $\Delta t$.
Finally, let $\BNM$, $\BMN$, and $\AMN$ be coefficients with dimensions such that
\begin{equation*}
\frac{N_{n\rightarrow m}}{\Delta t}
=\underset{\ABSORPTION}{\BNM N_n \RHO},
\qquad
\frac{N_{m\rightarrow n}}{\Delta t}
=\underset{\INDUCED}{\BMN N_m \RHO}
+
\underset{\SPONTANEOUS}{\AMN N_m}
\end{equation*}

\noindent
Absorption and induced emission are proportional to $\RHO$
which is the radiant energy density of the gas
as a function of radiant frequency $\nu$ and temperature $T$.
The $A$ and $B$ coefficients do not depend on $T$.

\bigskip
\noindent
At equilibrium, the transition rates are equal.
\begin{equation*}
\frac{N_{n\rightarrow m}}{\Delta t}=\frac{N_{m\rightarrow n}}{\Delta t}
\end{equation*}

\noindent
Hence
\begin{equation*}
\underset{\ABSORPTION}{\BNM N_n \RHO}
=\underset{\INDUCED}{\BMN N_m \RHO}
+\underset{\SPONTANEOUS}{\AMN N_m}
\end{equation*}

\noindent
Divide through by $N$.
\begin{equation*}
\underset{\ABSORPTION}{\BNM \frac{N_n}{N} \RHO}
=\underset{\INDUCED}{\BMN \frac{N_m}{N} \RHO}
+\underset{\SPONTANEOUS}{\AMN \frac{N_m}{N}}
\end{equation*}

\noindent
Then by the Maxwell-Boltzmann distribution (1) we have
\begin{equation*}
\underset{\ABSORPTION}{\BNM p_n \RHO\exp\left(-\frac{E_n}{kT}\right)}
=\underset{\INDUCED}{\BMN p_m \RHO\exp\left(-\frac{E_m}{kT}\right)}
+\underset{\SPONTANEOUS}{\AMN p_m\exp\left(-\frac{E_m}{kT}\right)}
\tag{2}
\end{equation*}

\noindent
Multiply both sides by $\exp(E_m/kT)$.
\begin{equation*}
\underset{\ABSORPTION}{\BNM p_n \RHO\exp\left(\frac{E_m-E_n}{kT}\right)}
=\underset{\INDUCED}{\BMN p_m \RHO}
+\underset{\SPONTANEOUS}{\AMN p_m}
\end{equation*}

\noindent
Note that for increasing $T$ we have
\begin{equation*}
\lim_{T\rightarrow\infty}\exp\left(\frac{E_m-E_n}{kT}\right)=1
\end{equation*}

\noindent
It follows that for $T\rightarrow\infty$ the equilibrium formula is
\begin{equation*}
\BNM p_n \RHO
=\BMN p_m \RHO
+\AMN p_m
\end{equation*}

\noindent
Divide through by $\RHO$.
\begin{equation*}
\BNM p_n=\BMN p_m+\frac{\AMN p_m}{\RHO}
\end{equation*}

\noindent
Energy density $\RHO$ increases with temperature $T$
hence $\AMN p_m/\RHO$ vanishes for $T\rightarrow\infty$ leaving
\begin{equation*}
\BNM p_n=\BMN p_m
\tag{3}
\end{equation*}

\noindent
Einstein reasoned that equation (3) is true in general based on the assumption that
the factors involved do not depend on $T$.
By substitution in the absorption term we can now eliminate $\BNM p_n$ and obtain
\begin{equation*}
\underset{\ABSORPTION}{\BMN p_m \RHO\exp\left(\frac{E_m-E_n}{kT}\right)}
=\underset{\INDUCED}{\BMN p_m \RHO}
+\underset{\SPONTANEOUS}{\AMN p_m}
\end{equation*}

\noindent
Divide both sides by $\BMN p_m$.
\begin{equation*}
\underset{\ABSORPTION}{\RHO\exp\left(\frac{E_m-E_n}{kT}\right)}
=\underset{\INDUCED}{\RHO}+\underset{\SPONTANEOUS}{\frac{\AMN}{\BMN}}
\end{equation*}

\noindent
Rearrange terms.
\begin{equation*}
\underset{\ABSORPTION}{\RHO\exp\left(\frac{E_m-E_n}{kT}\right)}
-\underset{\INDUCED}{\RHO}=\underset{\SPONTANEOUS}{\frac{\AMN}{\BMN}}
\end{equation*}

\noindent
Factor out $\RHO$.
\begin{equation*}
\RHO\left(\exp\left(\frac{E_m-E_n}{kT}\right)-1\right)
=\frac{\AMN}{\BMN}
\end{equation*}

\noindent
Solve for $\RHO$.
\begin{equation*}
\RHO
=\frac{\AMN}{\BMN}\,\frac{1}{\exp\left(\frac{E_m-E_n}{kT}\right)-1}
\end{equation*}

\noindent
We now consider an approximation of $\RHO$ for the case of a large exponential term.
\begin{equation*}
\RHO\approx
\frac{\AMN}{\BMN}\exp\left(-\frac{E_m-E_n}{kT}\right)
\end{equation*}

\noindent
Then by equivalence with Wien's law (which is accurate for large $\nu$)
\begin{equation*}
\rho_\text{wien}(\nu,T)=\frac{2h\nu^3}{c^2}\exp\left(-\frac{h\nu}{kT}\right)
\end{equation*}
we have
\begin{equation*}
\frac{\AMN}{\BMN}=\frac{2h\nu^3}{c^2}
\tag{4}
\end{equation*}
and
\begin{equation*}
E_m-E_n=h\nu
\end{equation*}

\noindent
Then by substitution we obtain Planck's law.
\begin{align*}
\RHO
&=\frac{\AMN}{\BMN}\,\frac{1}{\exp\left(\frac{E_m-E_n}{kT}\right)-1}
\\[1.5ex]
&=\frac{2h\nu^3}{c^2}\,\frac{1}{\exp\left(\frac{h\nu}{kT}\right)-1}
\end{align*}

\noindent
Let us now consider the values of the $A$ and $B$ coefficients.
The coefficient for spontaneous emission can be computed from quantum mechanics.
For example, for the hydrogen transition $2p\leftrightarrow1s$ we have
\begin{equation*}
A_{21}=\frac{e^{10}m_e}{26244\,\pi^5\varepsilon_0^5\hbar^6 c^3}
=6.27\times10^8\,\text{second}^{-1}
\end{equation*}

\noindent
The coefficient for induced emission can be obtained from equation (4).
\begin{equation*}
\BMN=\frac{c^2}{2h\nu^3}\,\AMN
\end{equation*}

\noindent
The coefficient for absorption can be computed from equation (3).
\begin{equation*}
\BNM=\frac{p_m}{p_n}\,\BMN
\end{equation*}

\noindent
The ratio $p_m/p_n$ is equal to $g_m/g_n$ where $g$ is the spin and orbital multiplicity
\begin{equation*}
g_n=(2S+1)(2\ell+1)
\end{equation*}
for quantum numbers $n$, $S$, and $\ell$.

\bigskip
\noindent
Hence for the transition $2p\leftrightarrow1s$ we have
\begin{align*}
g_1&=2\qquad(n=1,\,S=\tfrac{1}{2},\,\ell=0)
\\
g_2&=6\qquad(n=2,\,S=\tfrac{1}{2},\,\ell=1)
\end{align*}

\end{document}
