\documentclass[12pt]{article}
\usepackage{amsmath}

\parindent=0pt

\begin{document}

Feynman and Hibbs problem 3-9

\bigskip
Find the kernel for a particle in a constant field $f$
where the Lagrangian is
\begin{equation*}
L=\frac{m}{2}\dot x^2+fx
\end{equation*}

From the $L$ given above we have
\begin{equation*}
\frac{d}{dt}\left(\frac{\partial L}{\partial\dot x}\right)=m\ddot x
\end{equation*}
and
\begin{equation*}
\frac{\partial L}{\partial x}=f
\end{equation*}

By equation (2.7) which is
\begin{equation*}
\frac{d}{dt}\left(\frac{\partial L}{\partial\dot x}\right)=\frac{\partial L}{\partial x}
\end{equation*}
we have
\begin{equation*}
\ddot x=\frac{f}{m}
\end{equation*}

Hence velocity is constant. It follows that $x(t)$ is velocity times time.
\begin{equation*}
x(t)=\left(\frac{x_b-x_a}{t_b-t_a}\right)(t-t_a)+x_a
\end{equation*}

By equation (2.1) which is
\begin{equation*}
S=\int_{t_a}^{t_b} L\,dt
\end{equation*}
we have
\begin{equation*}
S_{cl}=\int_{t_a}^{t_b}\frac{m}{2}\dot x(t)^2\,dt
+\int_{t_a}^{t_b}fx(t)\,dt
\end{equation*}

Equation (2.8) provides the following solution to the first integral.
\begin{equation*}
\int_{t_a}^{t_b}\frac{m}{2}\dot x(t)^2\,dt
=\frac{m(x_b-x_a)^2}{2(t_b-t_a)}
\tag{1}
\end{equation*}


Solve the second integral.
\begin{equation*}
\int_{t_a}^{t_b}fx(t)\,dt=\frac{(x_b+x_a)(t_b-t_a)f}{2}
\tag{2}
\end{equation*}


Action $S_{cl}$ is the sum of (1) and (2).
\begin{equation*}
S_{cl}=\frac{m(x_b-x_a)^2}{2(t_b-t_a)}+\frac{(x_b+x_a)(t_b-t_a)f}{2}
\end{equation*}


By equation (3.51) which is
\begin{equation*}
K(b,a)=F(t_b-t_a)\exp\left(\frac{iS_{cl}}{\hbar}\right)
\end{equation*}
we have
\begin{equation*}
K(b,a)=F(t_b-t_a)\exp\left(\frac{im(x_b-x_a)^2}{2\hbar(t_b-t_a)}+\frac{i(x_b+x_a)(t_b-t_a)f}{2\hbar}\right)
\tag{3}
\end{equation*}


We now proceed to compute $F$. By equation (2.31) which is
\begin{equation*}
K(b,a)=\int_{-\infty}^\infty K(b,c)K(c,a)\,dx_c
\end{equation*}
we have
\begin{multline*}
K(b,a)=\int_{-\infty}^\infty
F(t_b-t_c)\exp\left(\frac{im(x_b-x_c)^2}{2\hbar(t_b-t_c)}+\frac{i(x_b+x_c)(t_b-t_c)f}{2\hbar}\right)
\\
{}\times F(t_c-t_a)\exp\left(\frac{im(x_c-x_a)^2}{2\hbar(t_c-t_a)}+\frac{i(x_c+x_a)(t_c-t_a)f}{2\hbar}\right)
\,dx_c
\tag{4}
\end{multline*}


Rewrite equation (4) as
\begin{equation*}
K(b,a)=F(t_b-t_c)F(t_c-t_a)\int_{-\infty}^\infty
\exp(iAx_c^2+iBx_c+iC)
\tag{5}
\end{equation*}
where
\begin{align*}
A&=\frac{m}{2\hbar}\left(\frac{1}{t_b-t_c}+\frac{1}{t_c-t_a}\right)
\tag{6}
\\
B&=\frac{(t_b-t_a)f}{2\hbar}-\frac{m}{\hbar}\left(\frac{x_b}{t_b-t_c}+\frac{x_a}{t_c-t_a}\right)
\tag{7}
\\
C&=\frac{f}{2\hbar}\big(x_b(t_b-t_c)+x_a(t_c-t_a)\big)+\frac{m}{2\hbar}
\left(\frac{x_b^2}{t_b-t_c}+\frac{x_a^2}{t_c-t_a}\right)
\tag{8}
\end{align*}


From the following formula
\begin{equation*}
\int_{-\infty}^{\infty}\exp(iAx_c^2+iBx_c+iC)\,dx_c
=\left(-\frac{\pi}{iA}\right)^{1/2}
\exp\left(-\frac{iB^2}{4A}+iC\right)
\end{equation*}
we have
{\footnotesize
\begin{multline*}
\int_{-\infty}^{\infty}\exp(iAx_c^2+iBx_c+iC)\,dx_c
=\left(-\frac{2\pi\hbar(t_b-t_c)(t_c-t_a)}{im(t_b-t_a)}\right)^{1/2}
\\
{}\times\exp\left(
\frac{im(x_b-x_a)^2}{2\hbar(t_b-t_a)}
+\frac{i(x_b+x_a)(t_b-t_a)f}{2\hbar}
-\frac{i(t_b-t_a)(t_b-t_c)(t_c-t_a)f^2}{8\hbar m}
\right)
\tag{9}
\end{multline*}
}


Substitute (9) into (5) to obtain
{\footnotesize
\begin{multline*}
K(b,a)=F(t_b-t_c)F(t_c-t_a)
\left(-\frac{2\pi\hbar(t_b-t_c)(t_c-t_a)}{im(t_b-t_a)}\right)^{1/2}
\\
{}\times \exp\left(
\frac{im(x_b-x_a)^2}{2\hbar(t_b-t_a)}
+\frac{i(x_b+x_a)(t_b-t_a)f}{2\hbar}
-\frac{i(t_b-t_a)(t_b-t_c)(t_c-t_a)f^2}{8\hbar m}
\right)
\tag{10}
\end{multline*}
}


Equating (3) with (10) reduces to
\begin{multline*}
F(t_b-t_a)=F(t_b-t_c)F(t_c-t_a)
\\
{}\times\left(-\frac{2\pi\hbar(t_b-t_c)(t_c-t_a)}{im(t_b-t_a)}\right)^{1/2}
\exp\left(-\frac{i(t_b-t_a)(t_b-t_c)(t_c-t_a)f^2}{8\hbar m}\right)
\tag{11}
\end{multline*}


From problem 3-7, let
\begin{equation*}
F(t)=\left(\frac{m}{2\pi i\hbar t}\right)^{1/2} g(t)
\tag{12}
\end{equation*}


Substitute (12) into (11) to obtain
\begin{align*}
&\left(\frac{m}{2\pi i\hbar(t_b-t_a)}\right)^{1/2} g(t_b-t_a)
\\
&\quad{}=\left(\frac{m}{2\pi i\hbar(t_b-t_c)}\right)^{1/2} g(t_b-t_c)
\left(\frac{m}{2\pi i\hbar(t_c-t_a)}\right)^{1/2} g(t_c-t_a)
\\
&\quad\quad{}\times\left(-\frac{2\pi\hbar(t_b-t_c)(t_c-t_a)}{im(t_b-t_a)}\right)^{1/2}
\exp\left(-\frac{i(t_b-t_a)(t_b-t_c)(t_c-t_a)f^2}{8\hbar m}\right)
\end{align*}


The coefficients cancel leaving
\begin{equation*}
g(t_b-t_a)=g(t_b-t_c)g(t_c-t_a)\exp\left(-\frac{i(t_b-t_a)(t_b-t_c)(t_c-t_a)f^2}{8\hbar m}\right)
\tag{13}
\end{equation*}


Note that
\begin{equation*}
(t_b-t_a)^3=
(t_b-t_c)^3
+3(t_b-t_a)(t_b-t_c)(t_c-t_a)
+(t_c-t_a)^3
\tag{14}
\end{equation*}


Solve for $g(t)$ by completing the cube.
\begin{equation*}
g(t)=\exp\left(-\frac{it^3f^2}{24\hbar m}\right)
\tag{15}
\end{equation*}


Substitute (15) into (12).
\begin{equation*}
F(t)=\left(\frac{m}{2\pi i\hbar t}\right)^{1/2} \exp\left(-\frac{it^3f^2}{24\hbar m}\right)
\tag{16}
\end{equation*}


Substitute (16) into (3).
\begin{multline*}
K(b,a)=\left(\frac{m}{2\pi i\hbar(t_b-t_a)}\right)^{1/2}
\\
{}\times\exp\left(
\frac{im(x_b-x_a)^2}{2\hbar(t_b-t_a)}
+\frac{i(x_b+x_a)(t_b-t_a)f}{2\hbar}
-\frac{i(t_b-t_a)^3f^2}{24\hbar m}
\right)
\end{multline*}

\end{document}
