\documentclass[12pt]{article}
\usepackage[margin=2cm]{geometry}
\usepackage{amsmath}

\begin{document}

\noindent
Consider a system with the following eigenstates.
\begin{align*}
|0\rangle&=(\text{1 0 0 0})^\dag\qquad\text{no electrons}\\
|1\rangle&=(\text{0 1 0 0})^\dag\qquad\text{one electron in state $\phi_1$}\\
|2\rangle&=(\text{0 0 1 0})^\dag\qquad\text{one electron in state $\phi_2$}\\
|3\rangle&=(\text{0 0 0 1})^\dag\qquad\text{two electrons, one in state $\phi_1$, one in state $\phi_2$}
\end{align*}

\noindent
Then for the wavefunction basis
\begin{equation*}
\phi_n(x)=\sqrt{\frac{2}{L}}\sin\left(\frac{n\pi x}{L}\right)
\end{equation*}

\noindent
and for $L=10^{-9}$ meters we have
\begin{equation*}
\hat{E}=\begin{pmatrix}
0&0&0&0\\
0&0.38\,\text{eV}&0&0\\
0&0&1.50\,\text{eV}&0\\
0&0&0&6.55\,\text{eV}
\end{pmatrix}
\end{equation*}

\noindent
Let $|\xi\rangle$ be the state vector
\begin{equation*}
|\xi\rangle
=\frac{1}{2}|0\rangle+\frac{1}{2}|1\rangle+\frac{1}{2}|2\rangle+\frac{1}{2}|3\rangle
=\begin{pmatrix}1/2\\1/2\\1/2\\1/2\end{pmatrix}
\end{equation*}

\noindent
The expected energy is
\begin{equation*}
\langle\xi|\hat{E}|\xi\rangle
=\frac{0\,\text{eV}}{4}+\frac{0.38\,\text{eV}}{4}+\frac{1.50\,\text{eV}}{4}+\frac{6.55\,\text{eV}}{4}
=2.11\,\text{eV}
\end{equation*}

\noindent
For the system we are considering, the result of a single measurement is either
0~eV, 0.38~eV, 1.50~eV, or 6.55~eV.
%The probability of observing each eigenvalue is 0.25 when the system is in state $|\xi\rangle$.
The value 2.11 eV is the expected average across multiple measurements.
Recall that a measurement causes the system to exit state $|\xi\rangle$
and enter an eigenstate $|0\rangle$, $|1\rangle$, $|2\rangle$, or $|3\rangle$
corresponding to the measured eigenvalue.
The system must be put back in state $|\xi\rangle$ before the next measurement.

\bigskip
\noindent
To use a slot machine analogy, state $|\xi\rangle$ is like the wheels spinning.
Observing the system makes the wheels stop.
The stopped wheels are in an eigenstate $|0\rangle$, $|1\rangle$, $|2\rangle$, or $|3\rangle$.
Once they are stopped the wheels don't change, they remain in the same eigenstate.
You have to pull the lever to get the wheels spinning again.

\end{document}
