\input{preamble}

\section*{Penguin anomaly}

Excerpt from {\it `Penguin' Anomaly Hints at Missing Particles}.\footnote{
Wolchover, Natalie.
{\it `Penguin' Anomaly Hints at Missing Particles.}\\
www.quantamagazine.org/penguin-anomaly-at-large-hadron-collider-hints-at-missing-particles-20150320/}

\begin{quote}
Most unexpected bumps in data go away as more data accumulate, just as you might get seven heads in your first 10 coin tosses only to end up with a 50-50 ratio after many more tosses. But after tripling their original sample size and analyzing approximately 2,400 of the rare penguin decays, the LHCb scientists say the anomaly hasn't diminished. Instead, it has lingered at an estimated statistical significance of ``3.7 sigma'' which means it is just as unlikely for such a large fluctuation to happen randomly as it would be to get 69 heads in 100 coin tosses. Physicists require a 5-sigma deviation from their expectations, equivalent to flipping 75 heads in 100 tosses (the odds of which are less than one in a million), to claim the discovery of a real effect.
\end{quote}

Recall that the binomial mass function with $p=1/2$
is the probability of obtaining exactly
$k$ heads in $n$ tosses.
\begin{equation*}
f(k)=\binom{n}{k}p^k(1-p)^{n-k}
\end{equation*}

In Eigenmath, define the binomial mass function
and calculate the probability
of getting exactly 75 heads in 100 tosses.

{\color{blue}
\begin{verbatim}
f(k) = choose(n,k) p^k (1-p)^(n-k)
n = 100
p = 1/2
float(f(75))
\end{verbatim}}

$1.91314\times10^{-7}$

\bigskip
Hence the probability of getting exactly 75 heads in 100 tosses is indeed
less than one in a million.

\bigskip
The mean for 100 coin tosses is 50 and the variance is 25.
\begin{equation*}
\mu=np=50,\quad\sigma^2=np(1-p)=100\times\tfrac{1}{2}\times\tfrac{1}{2}=25
\end{equation*}

The standard deviation is $\sigma=5$.
Hence a 5-sigma deviation from the mean is
$50+5\sigma=75$ and a
3.7-sigma deviation is $50+3.7\sigma=68.5$.

\bigskip
We have found the probability of getting exactly 75 heads in 100 tosses.
Let us now find the combined probability of {\it any} unlikely result,
such as 76 heads or 25 heads, etc.
Start with the following cumulative distribution function.
\begin{equation*}
\Pr(X\le x)=F(x)=\sum_{k=0}^xf(k)
\end{equation*}

Define as ``unlikely'' any number of heads that is beyond 5 standard deviations of the
mean $\mu=50$.
Then the probability of {\it any} unlikely result after 100 tosses is
\begin{equation*}
\Pr(X\le25)+\Pr(X\ge75)=F(25)+1-F(74)=2\,F(25)
\end{equation*}

The last equivalence is by symmetry of the binomial function.

\bigskip
In Eigenmath, define the cumulative distribution function and compute the probability
of fewer than 26 or more than 74 heads in 100 tosses.

{\color{blue}
\begin{verbatim}
F(x) = sum(k,0,x,f(k))
float(2 F(25))
\end{verbatim}}

$5.63628\times10^{-7}$

\bigskip
The probability is still less than one in a million.

\end{document}
